High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2020-10-09 Draft_V4046Sgr_mnras.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[letters, usenatbib]{mnras}

@@

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols

@@

% Abstract of the paper
\begin{abstract}
    %context
    V4046 Sgr is a nearby pre-main sequence spectroscopic binary system that has a gas-rich circumbinary disc.
    %aim
    We report on new $\sim$20\,mas resolution Atacama Large Millimeter Array (ALMA) observations of the radio continuum at 1.3\,mm, and compare with archival SPHERE-IRDIS differential polarised imaging (DPI) of scattered-light at H-band (1.65\,$\micron$) to reach conclusions on the disc structure. 
    %methods
    We use a parametric a radiative transfer model of the circumbinary disc, computed with \textsc{radmc3d} to interpret the available data, including the spectral-energy-distribution, in terms of a physical model. In addition, we explore decompositions of the imaging data in polar coordinates to better constrain the disc radial structure. 
    %results
    We report a two ringed structure for the millimetre-size dust population with a narrow, previously unseen, less than 4.17$\pm$0.94\,au-wide inner ring centred at 13.46$\pm$0.43\,au. This ring is the finest reported so far in a gas rich protoplanetary disc. The ring is surrounded by a $\sim$9\,au-wide gap, flanked by a very sharp outer ring starting at 24\,au. The radial profile of the outer ring exhibits an intriguing discontinuity at $\sim$35\,au. The RT model reproduces accurately the available data. We predict a three ringed structure for the small-dust population with a inner ring at $\sim$6\,au hiding under the coronagraph in the SPHERE observations. The shadow of the secondary reported previously from the scattered-light data is also reproduced by the RT model. 
    %conclusions
    The DPI image reveals strong forward-scattering at 1.65\,$\micron$, which we reproduced with rather large dust grains of no less than 0.4\,$\micron$, hinting at a deficit of very-small-grains. Using the parametric model scale height as an estimated height, we found that the thin inner ring is stable. Deeper ALMA observations may confirm the lopsided 5\,au ring.

%briefly describe the aims, methods, and main results
%single paragraph not more than 250 words
%No references should appear
\end{abstract}

@@

Recent observations of circumstellar discs have revolutionised the way we think about protoplanetary discs, but the focus of resolved imaging with the Atacama Large Millimeter Array (ALMA) or with VLT/SPHERE has been mainly been towards the brighter sources. This report is part of a survey of discs around T-Tauri Stars (DARTSS - ref Avenhaus, Perez et al in pret). Here, we present new 1.3\,mm continuum ALMA observation of V4046\,Sgr.

V4046 Sagittarii (Sgr) is is a double-lined spectroscopic binary of a nearly equal-mass pair of K-type stars (K5 and K7) with masses of 0.90$\pm$0.05\,M$_{\sun}$ and 0.85$\pm$0.04\,M$_{\sun}$ \citep{Rosenfeld_2012} on a close ($a \approx 0.041$\,au), circular ($e\leq0.001$) orbit, with an orbital period of 2.42 days \citep{refId0}. The V4046\,Sgr system locate at 72.41$\pm$0.34\,pc \citep{Gaia}, and is a member of the $\beta$ Pictoris moving group \citep{Zuckerman_2004}, fixing its age at 23$\pm$3\,Myr \citep{Mamajek_2014}. V4046\,Sgr binary system is surrounded by a massive ($\sim$0.1\,M$_{\sun}$), gas-rich disc extending to $\sim$300\,au \citep{Rosenfeld_2013, Rodriguez_2010}. The disc has a vast variety of molecules with a detailed subarcsecond ALMA molecular line imaging study by \citet{Kastner_2018}. The chemical structure is composed of CO and HCN that display centrally peaked morphologies, a sequence of sharp and diffuse rings of complex nitrile group molecules (HC$_3$N, CH$_3$CN), deuterated molecules (DCN, DCO$^+$), hydrocarbons (C$_2$H) and others like N$_2$H$^+$, H$_2$CO.

To analyse the new ALMA data, the aim was to reproduce the disc through a parametric model computed with the radiative transfer code \textsc{radmc3d} \citep{Dullemond_2012} and thus be able to interpret the data in terms of a physical model. The model would have to agree not only with the new ALMA observation but also with a previous SPHERE-IRDIS observation and the Spectral Energy Distribution (SED). Combining the model representation of the dust density distribution with a decomposition in polar coordinates of the ALMA image we can have better approximations of the radial structure of the disc. It is worth mentioning that previous models of V4046\,Sgr have been made \citep{Ru_z_Rodr_guez_2019, Rosenfeld_2013, 2019ApJ...882..160Q} but with the difference that our new reference observation of radio wavelengths has a sub-arcsecond resolution, allowing us to discover previously unseen underlying substructures of the large-dust grain disc.

The structure of this letter is as follows. The new 1.3\,mm continuum observation is described alongside with a SPHERE-IRDIS \textit{H}-band image in Section \ref{sec:Observations}. The structure of the parametric model used to simulate the disc is presented in Section \ref{sec:model}. The results of the model and the discussion are commented Section \ref{sec:results} and the main conclusions are described in Section \ref{sec:Conclusions}.

@@

Disc structures can be explored by means of comparing mm continuum and scattered-light images. These observations allow us to see the distribution of different populations of dust particles, while ALMA traces the millimetre-sized grains settled in the mid-plane of the disc, scattered-light imaging accounts for photons reflected off small micron-sized dust at the disc surface layer, high above the mid-plane. By comparing both, is easier to determine whether asymmetries in scattered light correspond to surface ripples or deep structures intrinsic to the underlying density distribution.

New ALMA observations of V4046\,Sgr were obtained in 2017 as part of the Cycle 5 program 2017.1.01167.S.. These were taken in the context of a larger survey for 14 optically visible TTauri discs at $\sim$50\,mas resolution. The survey simultaneously mapped 1.3\,mm continuum and the J = 2$-$1 line of $^{12}$CO with the C43-8 array configuration in band 6 (211-275\,GHz). This work focuses exclusively on the data obtained in the 1.3\,mm continuum for our source, taken with a beam size of $0\farcs04 \,\times \, 0\farcs05$.

In this investigation a previous polarimetric image at 1.65\,$\micron$ is used. The image was taken at the ESO Very Large Telescope with the SPHERE-IRDIS instrument, a complete description of the image and all about data reduction was reported in \citet{Avenhaus_2018}.

For our analysis of the ALMA image we used an image reconstruction strategy, using the non-parametric image synthesis of the \textsc{uvmem} package \citep{2006ApJ...639..951C, 2018A&C....22...16C} it is possible to super-resolve the clean beam, obtaining an effective angular resolution $\sim$2-3 times higher than in the clean image with natural weights. The model obtains this higher resolution through fitting the data using Chi-squared minimisation. Here we adopted a pure $\chi ^2$ model image. The resulting \textsc{uvmem} image shown in Fig. \ref{fig:two} exhibits previously unseen substructure of the disc. This image reveals that the disc features two rings of large dust grains with a broad gap between them. The wide and brighter outer ring has its peak intensity at $\sim$30\,au, has a surface brightness break at $\sim$35\,au, and reaches out to $\sim$60\,au. The thin and weak inner ring had never seen before, however \citet{Ru_z_Rodr_guez_2019} did anticipate its existence.

\begin{figure*}
	\includegraphics[width=\textwidth]{hot_two_E.pdf}
    \caption{Comparison of observations and simulated images at 1.65\,$\micron$ and 1.3 mm continuum of the circumbinary disc orbiting V4046\,Sgr. From top to bottom: DPI image and ALMA Band 6  observations; simulated images of the parametric model. Top left panel: SPHERE-IRDIS \textit{H}-band image with a yellow filled circle that illustrate the N\_ALC\_YJH\_S coronagraph (inner working angle $\sim0\farcs12$, or $\sim$8.6\,au at 72.4\,pc). Top right panel: 1.3\,mm continuum \textsc{uvmem} model image. The yellow ellipse shows the size of the natural-weighted beam: $ 0\farcs04 \, \times \, 0\farcs05$. The inset displays a enlarged view of the central emission, the green star marks the star positions. Bottom left panel: 1.65\,$\upmu m$ simulated image of the parametric model. Bottom right panel: 1.3\,mm simulated image of the parametric model. The yellow ellipse shows the size of the simulated beam: $ 0\farcs04 \, \times \, 0\farcs05$. For all the images in the figure the colour scale is linear.}

    \label{fig:two}
\end{figure*}

@@

Interestingly, the ALMA images also detects faint the 1.3\,mm continuum near the stellar positions (See the inset in Fig. \ref{fig:two}). Since this faint central emission is offset from the star, at the origin of coordinates in the figure, it is probably due to large dust grains. This dust structure is at a distance of only $\sim0\farcs02$ from the binary system.

@@

For the representation of the stars, we used two Kurucz photospheres models \citep{1979ApJS...40....1K, 1997A&A...318..841C} with T$_{\mathrm{eff},1} =$ 4350\,K, R$_{*,1} =$ 1.064\,R$_{\sun}$, M$_{*,1} =$ 0.90\,M$_{\sun}$ and T$_{\mathrm{eff},2} =$ 4060\,K, R$_{*,2} =$ 1.033\,R$_{\sun}$, M$_{*,2} =$ 0.85\,M$_{\sun}$ respectively and with an accretion rate of log$\,\dot{\mathrm{M}} = -$9.3\,Myr$^{-1}$ for both cases \citep{10.1111/j.1365-2966.2011.19366.x}. The stars were placed with a separation of 0.041\,au and oriented with a 30\,$\degr$ rotation, relative to the semi-major axis of the disk, to mach the shadows reported by \citet{dOrazi}.

@@

Assuming a three dimensional model in a spherical reference frame with coordinates ($r$, $\theta$, $\phi$), gas density ($\rho_{\mathrm{gas}}$) distribution follows
\begin{equation}
  \rho_{\mathrm{gas}}(r,z) =\frac{\Sigma(r) \,\delta(r)}{\sqrt{2\pi} \,r H(r)}  \mathrm{exp}\left[-\frac{1}{2} \left(\frac{z}{r H(r)}\right)^2\right].
\end{equation}
Where $\delta(r)$ is a parameter that indicates the density drops in the gaps and cavities, $H(r)$ is the scale height profile and $\Sigma(r)$ been the surface density profile. $\Sigma(r)$ is defined by
\begin{equation}
  \Sigma(r) = \Sigma_\mathrm{c} \left(\frac{r}{R_\mathrm{c}}\right)^{-\gamma}  \, \mathrm{exp}\left[-\left(\frac{r}{R_\mathrm{c}}\right)^{2-\gamma}\right],
\end{equation}
where $R_c$ is a characteristic radius and $\gamma$ is the surface density gradient. A fixed $\gamma$ = 1 is used as it is a typical value for discs \citep{Andrews_2009,Andrews_2010}.

@@

The simulated image at 1.65\,$\micron$ shows a similar radial structure to the one visible in the observations, displaying a two ringed disc, where the innermost small-dust ring of the parametric model hides under the artificial coronagraph. The visible asymmetry in the SPHERE observations is recreated using grains larger than 0.4\,$\micron$ as smaller grains do not cause a strong forward scattering, meaning that the disc is depleted of very-small grains. Interestingly, the model accurately shows the shadows described by \citet{dOrazi} that are present in the Sphere-IRDIS image.

@@

The key conclusions of our analysis are as follows.
\begin{itemize}
  \item We report a narrow inner ring for the 1.3\,mm continuum located at 13.46$\pm$0.43\,au from the stars and has an estimated width of 4.17$\pm$0.94\,au. The location of this ring is similar to the inner ring observed in the scattered-light image, revealing that the ring includes a considerable mass of millimetre-sized grains of around 3\,M$_{\earth}$. Using the parametric model scale height value ($h= $ 0.28\,au at 13.46\,au) we have that the ring width is $\sim$15 times its estimated height, making it a stable ring.

@@

  \item As our parametric model accounts for the SED of the system, the disc requires the existence of a sub-micron dust population close (<5\,au) to the stars. We also predict the existence of another thin ring at $\sim$6\,au, about 3\,au-wide and made of small dust grains that lies under the coronagraph of the scattered-light image. Additionally, the weak central emission at 1.3\,mm could be part of this ring.

\end{itemize}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[letters,usenatbib,times]{mnras}

@@

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
%% \usepackage{amsmath}	% Advanced maths commands
%% \usepackage{amssymb}	% Extra maths symbols

@@

% Abstract of the paper
\begin{abstract}
    V4046 Sgr is a nearby pre-main sequence spectroscopic binary system that hosts a gas-rich disc known for its wide gaps and dusty ring as seen by the Atacama Large Millimeter/submillimeter Array (ALMA).    We present new high resolution ($\sim$20\, mas) ALMA observations of the 1.3\,mm radio continuum. These observations, combined with SPHERE-IRDIS polarized scattered light images, allow us to characterize the disc structure.  We reproduce the available observations, including the spectral energy distribution, in terms of a physical structure using a parametric radiative transfer (RT) model.  We report a two-ringed structure for the 1.3 mm dust emission with a narrow, previously unseen, less than 4.17$\pm$0.94\,au-wide inner ring centred at 13.46$\pm$0.43\,au.  The ring is surrounded by a $\sim$9\,au-wide gap, flanked by a very sharp outer ring starting at 24\,au.  We deproject the images in polar coordinates to better constrain the disc radial structure.  The radial profile of the outer ring exhibits an intriguing discontinuity at $\sim$35\,au.   The model predicts a three-ringed structure for the small-dust population with an inner ring at $\sim$6\,au hiding under the coronagraph in the SPHERE observations.  The shadow of the secondary star reported previously from scattered-light data is also reproduced by our RT model. The polarized light image reveals strong forward-scattering at 1.65\,$\micron$, which we reproduced with dust grains of no less than 0.4\,$\micron$, hinting at a deficit of very-small-grains. Using the parametric model scale-height as an estimated height, we found that the thin inner ring is stable. Deeper ALMA observations may confirm the lopsided 6\,au ring.
\end{abstract}

@@

Recent observations of circumstellar discs have revolutionised the way we think about protoplanetary discs, but the focus of resolved imaging with the Atacama Large Millimeter/submillimeter Array (ALMA) or with VLT/SPHERE has been mainly been towards the brighter sources. ***EXPAND, I would give a general description on substructures and cite ``e.g. Andrews (2020) for a recent review''***

V4046 Sagittarii (Sgr) is is a double-lined spectroscopic binary of K-type stars (K5 and K7) with very similar masses of 0.90$\pm$0.05\,M$_{\sun}$ and 0.85$\pm$0.04\,M$_{\sun}$ \citep{Rosenfeld_2012} on a close ($a \approx 0.041$\,au), circular ($e\leq0.001$) orbit, with an orbital period of 2.42 days \citep{refId0}. The V4046\,Sgr system, located at 72.41$\pm$0.34\,pc \citep{Gaia}, is a member of the $\beta$ Pictoris moving group \citep{Zuckerman_2004}, with an estimated age of 23$\pm$3\,Myr \citep{Mamajek_2014}. The V4046\,Sgr binary system hosts a massive ($\sim$0.1\,M$_{\sun}$) circumbinary disc extending to $\sim$300\,au \citep{Rosenfeld_2013, Rodriguez_2010}. The disc has a vast variety of molecules with a detailed sub-arcsecond ALMA molecular line imaging study by \citet{Kastner_2018}.

The Disks ARound T Tauri Stars (DARTTS) program was first presented in \citet{Avenhaus_2018} with the aim to image eight stars with the Spectro-Polarimeter High-contrast Exoplanet REsearch (SPHERE). The sample is not biased towards exceptionally bright and large disks, and consists of only solar-mass stars. A second part of the survey increased the number of sources by presenting 21 new images of circumstellar disks \citep{Garufi2020}.  The observations revealed diversed structures and morphologies in the scattering surface of these disks.  This letter is part of our new DARTTS survey with ALMA (DARTTS-A) which will present millimeter observations of nine protoplanetary disks previously imaged in polarized scattered light in DARTTS-S.

Here, we present new 1.3\,mm continuum ALMA observation of V4046\,Sgr. The structure of this letter is as follows. The new 1.3\,mm continuum observation is described alongside with a SPHERE-IRDIS \textit{H}-band image in Section \ref{sec:Observations}. The structure of the parametric model used to simulate the disc is presented in Section \ref{sec:model}. It is worth mentioning that previous models of V4046\,Sgr have been made \citep{Ru_z_Rodr_guez_2019, Rosenfeld_2013, 2019ApJ...882..160Q} but with the difference that our new reference observation of radio wavelengths has a sub-arcsecond resolution, allowing us to discover previously unseen underlying substructures of the large-dust grain disc.  The results of the model and the discussion are commented Section \ref{sec:results} and the main conclusions are described in Section \ref{sec:Conclusions}.

@@

%% Disc structures can be explored by means of comparing mm continuum and scattered-light images. These observations allow us to see the distribution of different populations of dust particles, while ALMA traces the millimetre-sized grains settled in the mid-plane of the disc, scattered-light imaging accounts for photons reflected off small micron-sized dust at the disc surface layer, high above the mid-plane. By comparing both, is easier to determine whether asymmetries in scattered light correspond to surface ripples or deep structures intrinsic to the underlying density distribution.

New ALMA observations of V4046\,Sgr were obtained in 2017 as part of the Cycle 5 program 2017.1.01167.S (PI: S. Perez). These were acquired in the context of a larger survey for 9 optically visible TTauri discs at $\sim$50\,mas resolution, the DARTTS-A program (Perez et al. {\em in prep}). The survey simultaneously mapped 1.3\,mm continuum and the J = 2$-$1 line of $^{12}$CO with the C43-8 array configuration in band 6 (211-275\,GHz). This work focuses exclusively on the data obtained in the 1.3\,mm continuum for our source, taken with a beam size of $0\farcs04 \,\times \, 0\farcs05$ (in natural weights).

In this investigation, we use the previous polarimetric image at 1.65\,$\micron$ part of the DARTTS-SPHERE I survey \citep{Avenhaus_2018}. The image was taken at the ESO Very Large Telescope with the SPHERE-IRDIS instrument in differential polarization imaging (DPI) mode, and a complete description of the data reduction was presented in \citet{Avenhaus_2018}.

For our analysis of the ALMA image we used an image reconstruction strategy, using the non-parametric image synthesis of the \textsc{uvmem} package \citep{2006ApJ...639..951C, 2018A&C....22...16C} it is possible to super-resolve the clean beam, obtaining an effective angular resolution $\sim$2-3 times higher than in the clean image with natural weights. The model obtains this higher resolution through fitting the data using Chi-squared minimisation. Here we adopted a pure $\chi^2$ model image. The resulting \textsc{uvmem} image shown in Fig. \ref{fig:two} exhibits previously unseen substructure of the disc. This image reveals that the disc features two rings of large dust grains with a broad gap between them. The wide and brighter outer ring has its peak intensity at $\sim$30\,au, has a surface brightness break at $\sim$35\,au, and reaches out to $\sim$60\,au. The thin and weak inner ring had never seen before, however \citet{Ru_z_Rodr_guez_2019} did anticipate its existence.

\begin{figure*}
  \includegraphics[width=\textwidth]{hot_two_E.pdf}
  \caption{Comparison of observations and simulated images at 1.65\,$\micron$ and 1.3 mm continuum of the circumbinary disc orbiting V4046\,Sgr. From top to bottom: DPI image and ALMA Band 6  observations; simulated images of the parametric model. Top left panel: SPHERE-IRDIS \textit{H}-band image with a yellow filled circle that illustrate the N\_ALC\_YJH\_S coronagraph (inner working angle $\sim0\farcs12$, or $\sim$8.6\,au at 72.4\,pc). Top right panel: 1.3\,mm continuum \textsc{uvmem} model image. The yellow ellipse shows the size of the natural-weighted beam: $ 0\farcs04 \, \times \, 0\farcs05$. The inset displays a enlarged view of the central emission, the green star marks the star positions. Bottom left panel: 1.65\,$\upmu m$ simulated image of the parametric model. Bottom right panel: 1.3\,mm simulated image of the parametric model. The yellow ellipse shows the size of the simulated beam: $ 0\farcs04 \, \times \, 0\farcs05$. For all the images in the figure the colour scale is linear.}
  \label{fig:two}
\end{figure*}

@@

Interestingly, the ALMA image also detects faint 1.3\,mm continuum emission near the stellar positions (See the inset in Fig. \ref{fig:two}). Since this faint central emission is offset from the star, at the origin of coordinates in the figure, it is probably due to large dust grains. This dust structure is at a distance of only $\sim0\farcs02$ from the binary system.

@@

For the representation of the stars, we used two Kurucz photospheres models \citep{1979ApJS...40....1K, 1997A&A...318..841C} with T$_{\mathrm{eff},1} =$ 4350\,K, R$_{*,1} =$ 1.064\,R$_{\sun}$, M$_{*,1} =$ 0.90\,M$_{\sun}$ and T$_{\mathrm{eff},2} =$ 4060\,K, R$_{*,2} =$ 1.033\,R$_{\sun}$, M$_{*,2} =$ 0.85\,M$_{\sun}$ respectively and with an accretion rate of log$\,\dot{\mathrm{M}} = -$9.3\,Myr$^{-1}$ for both cases \citep{10.1111/j.1365-2966.2011.19366.x}. The stars were placed with a separation of 0.041\,au and oriented with a 30\,$\degr$ rotation, relative to the semi-major axis of the disk, to match the shadows reported by \citet{dOrazi}.

@@

Assuming a three dimensional model in a spherical reference frame with coordinates ($r$, $\theta$, $\phi$), the gas density ($\rho_{\mathrm{gas}}$) distribution follows
\begin{equation}
  \rho_{\mathrm{gas}}(r,z) =\frac{\Sigma(r) \,\delta(r)}{\sqrt{2\pi} \,r H(r)}  \mathrm{exp}\left[-\frac{1}{2} \left(\frac{z}{r H(r)}\right)^2\right].
\end{equation}
Where $\delta(r)$ is a parameter that indicates the density drops in the gaps and cavities, $H(r)$ is the scale height profile and $\Sigma(r)$ been the surface density profile. $\Sigma(r)$ is defined by
\begin{equation}
  \Sigma(r) = \Sigma_\mathrm{c} \left(\frac{r}{R_\mathrm{c}}\right)^{-\gamma}  \, \mathrm{exp}\left[-\left(\frac{r}{R_\mathrm{c}}\right)^{2-\gamma}\right],
\end{equation}
where $R_c$ is a characteristic radius and $\gamma$ is the surface density gradient. A fixed $\gamma$ = 1 is used as it is a typical value for discs \citep{Andrews_2009,Andrews_2010}.

@@

The simulated image at 1.65\,$\micron$ shows a similar radial structure to the one visible in the observations, displaying a two ringed disc, where the innermost small-dust ring of the parametric model hides under the artificial coronagraph. The visible asymmetry in the SPHERE observations is recreated using grains larger than 0.4\,$\micron$ as smaller grains do not cause a strong forward scattering, meaning that the disc is depleted of very-small grains. Interestingly, the model accurately shows the shadows described by \citet{dOrazi} that are present in the SPHERE-IRDIS image.

@@

The key conclusions of our analysis are as follows.
\begin{enumerate}

  \item We report a narrow inner ring for the 1.3\,mm continuum located at 13.46$\pm$0.43\,au from the stars and has an estimated width of 4.17$\pm$0.94\,au. The location of this ring is similar to the inner ring observed in the scattered-light image, revealing that the ring includes a considerable mass of millimetre-sized grains of around 3\,M$_{\earth}$. Using the parametric model scale height value ($h= $ 0.28\,au at 13.46\,au) we have that the ring width is $\sim$15 times its estimated height, making it a stable ring.

@@

  \item As our parametric model accounts for the SED of the system, the disc requires the existence of a sub-micron dust population close (<5\,au) to the stars. We also predict the existence of another thin ring at $\sim$6\,au, about 3\,au-wide and made of small dust grains that lies under the coronagraph of the scattered-light image. Additionally, the weak central emission at 1.3\,mm could be part of this ring.
\end{enumerate}
